---
title: "Simulate Synthetic Count Data"
author: "Mukai Wang"
date: "2025-03-19"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulate Synthetic Count Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message=FALSE, echo=T)
options(mc.cores = 4)
```


In this vignette I will demonstrate how to fit four distributions (poisson, negative binomial, Gamma and log normal) with or without zero inflation.

# Load example data

First load the example data from [Ascinar etal](https://doi.org/10.1038/s41591-020-01183-8). This `SummarizedExperiment` object includes 1098 gut microbiome samples. We extract out the count matrix. Note that the rows are features and the columns are samples. I already filtered out features whose prevalence is smaller than 20%.


```{r}
library(simlite)
data("gut_experiment_subset", package="simlite")
print(dim(gut_experiment_subset))
count_data <- SummarizedExperiment::assay(gut_experiment_subset)
```


We normalize the count matrix so that the library size is the same for all the samples

```{r}
depths <- colSums(count_data)
min_depth <- min(depths)
normalization_factor <- min_depth / t(replicate(nrow(count_data), depths))
count_data_normalized <- count_data * normalization_factor
```


There are many rare features which makes fitting distributions difficult. Let me focus on features which are observed in at least 50% of samples.


```{r}
count_data_normalized <- round(count_data_normalized)
prevalences <- rowMeans(count_data_normalized > 0)
subset_features <- which(prevalences > 0.5)
# this step is used to make the vignette run faster
subset_features <- sample(subset_features, 100)
count_data_subset <- count_data_normalized[subset_features, ]
```


# Fit Distributions

Fit Poisson distribution.

```{r}
pois_params <- fitdistr_mat(count_mat=count_data_subset, dist="pois", cutoff=0.97, ncores=4)
head(pois_params)
```
Fit zero-inflated Poisson distribution.


```{r}
zipois_params <- fitdistr_mat(count_mat=count_data_subset, dist="zipois", cutoff=0.97, ncores=4)
head(zipois_params)
```

Fit negative binomial distribution.

```{r}
nb_params <- fitdistr_mat(count_mat=count_data_subset, dist="nb", cutoff=0.97, ncores=4)
head(nb_params)
```



Fit zero inflated negative binomial 


```{r}
zinb_params <- fitdistr_mat(count_mat=count_data_subset, dist="zinb", cutoff=0.97, ncores=4)
head(zinb_params)
```

Fit zero-inflated Gamma distribution.


```{r}
zigamma_params <- fitdistr_mat(count_mat=count_data_subset, dist="zigamma", cutoff=0.97, ncores=4)
head(zigamma_params)
```


Fit zero-inflated lognormal distributions.


```{r}
zilnorm_params <- fitdistr_mat(count_mat=count_data_subset, dist="zilnorm", cutoff=0.97, ncores=4)
head(zilnorm_params)
```




# Simulate Counts


For now I just illustrate simulations where features are not correlated. First generate a copula with the expected number of samples and number of features.

```{r}

nsample <- 30
nfeature <- 60
copula <- gausscopula(n=nsample, Sigma=diag(nrow=nfeature))
pois_counts <- simcountmat(copula=copula, params=pois_params[1:nfeature, ,drop=F],
                           dist="pois")

```



```{r}

zipois_counts <- simcountmat(copula=copula, params=zipois_params[1:nfeature, ,drop=F],
                           dist="zipois", ncores=4)

nb_counts <- simcountmat(copula=copula, params=nb_params[1:nfeature, ,drop=F],
                           dist="nb", ncores=4)

zinb_counts <- simcountmat(copula=copula, params=zinb_params[1:nfeature, ,drop=F],
                           dist="zinb", ncores=4)

```



```{r}

zigamma_counts <- simcountmat(copula=copula, params=zigamma_params[1:nfeature, ,drop=F],
                           dist="zigamma", ncores=4)

zilnorm_counts <- simcountmat(copula=copula, params=zilnorm_params[1:nfeature, ,drop=F],
                           dist="zilnorm", ncores=4)

```


